---
title: "Voting & Income Final Report"
author: "Kendra Lyons"
date: "12/10/2021"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
library(effects)
library(ggplot2)
library(ggthemes)
library(knitr)
library(tidyverse)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Recent Census and polling data confirms that there is a disparity in voting between 
low-income and high-income citizens. There is a clear positive relationship between 
voting and income. Simply put, people with less money are less inclined to vote, 
and vice versa. The same data can give us insight into why this is the case, since 
the U.S. Census Bureau and polling organizations like FiveThirtyEight include 
questions about why people don't vote, as well as how they feel about political 
institutions in general. Analysis of such data is invaluable for understanding what
prevents lower income people from voting and how we as a society can enact policy
changes that make democratic participation more accessible to people of all income
levels. The primary research questions guiding this project are: 1) How do voting 
patterns differ according to income level? and 2) What factors prevent low income 
citizens from voting as often as high income citizens do?

# Data

Data from two sources, the U.S. Census Bureau and FiveThirtyEight, were central to 
this project. Table 7, entitled "Reported Voting and Registration of Family Members, 
by Age and Family Income", and Table 10, entitled "Reasons For Not Voting, By Selected 
Characteristics" from the U.S. Census Bureau's 2020 Current Population Survey were
both used, and can be freely accessed on their website at: <https://www.census.gov/data/tables/time-series/demo/voting-and-registration/p20-585.html>.
From Table 7, only the data on all adults (rows 8-18) was selected, since, at this 
time, how age influences voting and registration is not being studied. Rather, this 
project is primarily focused on how household income affects voting and registration. 
For the same reason, only the data from Table 10 that is broken down by income (rows 
36-47) was selected for analysis. Census data was aggregated by income bracket and 
both raw numbers (in thousands) and percentages were included. Respondents were asked 
whether they were registered to vote, whether they voted, and if they reported not 
voting, what the reason was. 

Data from a FiveThirtyEight & Ipsos poll conducted in September 2020 was obtained 
from their GitHub repository "non-voters", located at: <https://github.com/fivethirtyeight/data/tree/master/non-voters>. These data were 
collected and used for a story published on October 26, 2020 entitled "Why Many 
Americans Don't Vote". Specifically, questions analyzed asked people whether 
"traditional parties and politicians" cared about people like them, how easy they 
considered voting in national elections to be, the importance of the results of the 
2020 presidential election, how frequently they vote and more. Data were also 
collected on income, education, voting frequency and more. According to the readme
attached to the Git repository, these data came from "5,239 respondents who matched 
to the voter file and whose verified vote history we have, and 597 respondents who 
did not match to the voter file and described themselves as voting "rarely" or 
"never," all of whom have been eligible for at least 4 elections"

This researcher’s goals in selecting this data were to explore and highlight the 
relationships between income and election participation, but education is also taken
into account. Income and Education can be seen as to aspects of class. This is 
important from a policy standpoint because understanding why voters do and don't vote 
and what socioeconomic factors may influence their decisions can help activists, 
canvassers, advocacy organizations, political parties and governments increase 
voter turnout. Identifying demographics with lower levels of voting can also encourage
policymakers to craft policy that benefits such populations, which is likely to 
encourage their democratic participation and personal investment in our political system.

For more information about the data, please refer to the following:

* <https://github.com/fivethirtyeight/data/tree/master/non-voters#readme>
* <https://www.census.gov/topics/public-sector/voting/about.html>

# Visualization

## Census Data

Read in and inspect data.

```{r}
voters_by_income <- read_csv("data/census_voters_all_incomes_tidy.csv")
glimpse(voters_by_income)

all_nonvoters <- read_csv("data/all_nonvoter_reasons_tidy.csv")
glimpse(all_nonvoters)

nonvoters_by_income <- read_csv("data/nonvoter_reasons_by_income_tidy.csv")
glimpse(nonvoters_by_income)

```

### Reported Voting by Income

This plot displays the striking correlation between voting and income in 2020. Its purpose is 
to demonstrate the trend of higher income correlating with greater likelihood of voting, and 
vice versa. The subsequent plots will dive deeper into the reasons people at different income
levels say they didn't vote in 2020 to contest characterizations of poor voters not voting due 
to apathy or laziness. to the contrary, they are much more likely to say they Nearly 85% of 
people with annual incomes of $150,000 or more voted in 2020, while just over 41% of people 
with annual incomes under $10,000 did. 

```{r}
# make palette
palette10 <- c("#d9ed92","#b5e48c","#99d98c","#76c893","#52b69a",
               "#34a0a4","#168aad","#1a759f","#1e6091","#184e77")
# barplot of percentage of (reported) voters by family income
voters_by_income %>%
  ggplot(aes(x=voted_percent, y=reorder(family_income, voted_percent)))+
  geom_col(aes(fill=reorder(family_income, voted_percent)), show.legend = FALSE)+
  geom_label(aes(label=voted_percent))+
  labs(title="Reported Voting by Income", 
       subtitle="November 2020 U.S. Census Data",
       caption="Source: U.S. Census Bureau, Current Population Survey, November 2020")+
  xlab("Percent of Citizens")+
  ylab("Family Income")+ 
  scale_fill_manual(values=palette10) +
  theme_dark()
```

*Figure 1*

### Nonvoter Reasons, All Incomes

This plot gives an overview of the general non-voting (registered) population in 2020 to explore what reasons people gave for not voting, regardless of income. This plot is meant to get people thinking about why people don't vote, and subsequent plots will dive deeper into the differences in reasons given by nonvoters with different incomes. Overall, lack of interest, not liking candidates or campaign issues, busy schedules and illness or disability are the most common reported reasons for not voting.

```{r}
palette12 <- c("#001219","#005f73","#0a9396","#94d2bd","#e9d8a6","#ee9b00","#ca6702",
                 "#bb3e03","#ae2012","#9b2226", "#5c110a", "#370E44")
all_nonvoters %>%
ggplot(aes(x = percent,
           y = reorder(reason, percent)))+
  geom_col(aes(fill = reorder(reason, percent)), show.legend = FALSE)+
  labs(title="Reported Reasons for Not Voting", 
       subtitle = "Registered Non-Voters with Reported Income", 
       x="Percent of Respondants", y="", 
       caption="Source: U.S. Census Bureau, Current Population Survey, November 2020")+
  geom_label(aes(label=percent)) +
  scale_fill_manual(values = palette12) +
  theme_dark()

```

*Figure 2*

### Nonvoter Reasons by Income

This plot breaks down reported reasons for not voting by income, demonstrating that 
the lowest income voters are more likely not to vote due to barriers like disability 
or illness, transportation issues or inconvenient polling locations. In contrast, 
higher income voters are more likely not to vote due to busy schedules, travel or not 
liking candidates or campaign issues. Disinterest or not liking candidates or campaign 
issues are reasons commonly given by people at all income levels, suggesting a broad 
crisis of faith in the U.S. political system. Notably, disability or illness and 
conflicting schedules were also among the top reasons for not voting overall. Both of 
these can be ameliorated in the future by providing early and mail-in ballot access to 
people of all income levels, and by making polling places much more accessible to low 
income people, particularly those with disabilities and transportation issues. 

```{r}
# factor income levels
nonvoters_by_income$income <- factor(nonvoters_by_income$income, 
                                     levels=c("<10K", "10-15K", "15-20K", "20-30K", "30-40K", 
                                              "40-50K", "50-75K", "75-100K", "100-150K", ">150K"))

nonvoters_by_income %>%
  ggplot(aes(x=percent, y=income))+
  geom_col(aes(fill = income), show.legend = FALSE)+
  facet_wrap(~reason) +
  labs(title="Reported Reasons for Not Voting by Income", 
       subtitle = "Registered Non-Voters with Reported Income", 
       x="Percent of Respondants in Income Bracket", y="Income", 
       caption="Source: U.S. Census Bureau, Current Population Survey, November 2020") +
  scale_fill_manual(values = palette10) +
  theme_dark()
```

*Figure 3*

## FiveThirtyEight Data

Read in and inspect data.
```{r}
voter_survey_538 <- read_csv("data/voter_survey_538_tidy.csv")

glimpse(voter_survey_538)
```

Factor income, pols_dont_care, voter_category and education.

```{r}
voter_survey_538$income_cat <- factor(voter_survey_538$income_cat, 
                                      levels = c("Less than $40k", "$40-75k", 
                                                 "$75-125k", "$125k or more"))


voter_survey_538$pols_dont_care <- factor(voter_survey_538$pols_dont_care, 
                                       levels = c("Strongly Disagree", "Somewhat Disagree", 
                                                  "Somewhat Agree", "Strongly Agree"))


voter_survey_538$voter_category <- factor(voter_survey_538$voter_category, 
                                      levels = c("always", "sporadic", "rarely/never"))


voter_survey_538$educ <- factor(voter_survey_538$educ, 
                                      levels = c("High school or less", "Some college",
                                                 "College"))
```

### Voting Frequency and Income

```{r}
vote_income_category <- voter_survey_538 %>%
  count(income_cat, voter_category) %>%
  group_by(income_cat) %>%
  mutate(income_percent = (n/sum(n))*100)

palette3 <- c("#2a9d8f", "#e9c46a", "#e76f51")

vote_income_category %>%
  filter(income_cat != "NA") %>%
  ggplot(aes(y = income_cat,
             x = income_percent)) +
  geom_col(aes(fill = voter_category)) +
  labs(title = "Voting Frequency and Income",
       x = '', y = 'Income', caption = "Data from FiveThirtyEight") +
  theme_linedraw() +
  scale_fill_manual(values = palette3) +
  theme(text = element_text(size = 16, face = "bold"),
        legend.title = element_blank(),
        legend.position = "top") +
  scale_x_reverse()
```

*Figure 4*

### Perception of Election Result Importance and Income

```{r}
# visualize "election results matter" by income
erm_by_income <- voter_survey_538  %>%
  filter(election_results_matter != "NA") %>%
  count(election_results_matter, income_cat) %>%
  group_by(income_cat) %>%
  mutate(income_cat_perc = (n/sum(n))*100)

palette2 <- c("#f4a261", "#2a9d8f")

erm_by_income %>%
  ggplot(aes(y = income_cat,
             x = income_cat_perc)) +
  geom_col(aes(fill = election_results_matter)) +
  theme_linedraw() +
  labs(title = "Election Result Importance and Income",
       x = '', y = 'Income', caption = "Data from FiveThirtyEight") +
  theme(text = element_text(size = 16, face = "bold"),
        legend.title = element_blank(),
        legend.position = "top") +
  guides(fill = guide_legend(ncol = 1)) +
  scale_x_reverse()  +
  scale_fill_manual(values = palette2)
```

*Figure 5*

### Perception of Importance to Political Entities and Income

```{r}

pdc_by_income <- voter_survey_538 %>%
  filter(pols_dont_care != "NA") %>%
  count(pols_dont_care, income_cat) %>%
  group_by(income_cat) %>%
  mutate(income_cat_perc = (n/sum(n))*100)

palette4 <- c("#e76f51","#f4a261","#e9c46a","#2a9d8f")

pdc_by_income %>%
  ggplot(aes(y = income_cat, 
             x = income_cat_perc)) +
  geom_col(aes(fill = factor(pols_dont_care))) +
  theme_linedraw() +
  labs(title = '"Traditional parties and politicians', 
       subtitle = 'don’t care about people like me."',
       x = '', y = 'Income', caption = "Data from FiveThirtyEight") +
  scale_fill_manual(values = palette4) +
  theme(text = element_text(size = 16, face = "bold"),
        legend.title = element_blank(),
        legend.position = "top") +
  guides(fill = guide_legend(ncol = 2))+
  scale_x_reverse()
```

*Figure 6*

# Modeling

## Visualization

Visualize relationship between income level, voting frequency and education.

```{r}
v_labels <- c("Rarely/Never", "Sporadic", "Always") 
i_labels <- c("Less than $40k", "$40-75k", "$75-125k", "$125k or more")

voter_survey_538 %>%
  ggplot(aes(x = income_cat_coded,
             y = voter_cat_coded,
             color = educ)) +
  geom_jitter(alpha = .5) +
  geom_smooth(method = lm) +
  theme(text = element_text(size = 16, face = "bold")) +
  labs(y = "Voter Category: Rarely/Never, Sporadic, Always",
       x = "Income: Less than $40k, $40-75k, $75-125k, $125k or more",
       title = "Income and Voting Frequency",
       caption = "Data from FiveThirtyEight") +
  scale_color_colorblind() +
  theme_bw()
```

*Figure 7*

Most respondants to the FiveThirtyEight poll appear to be classified as sporadic voters.
FiveThirtyEight categorized respondents using the voting records matched to them. According
to their Non-voters Codebook, sporadic voters "voted in at least two, but fewer than 
all-but-one of the elections they were eligible in." People with yearly income less than 
$40,000 were more likely to be classified as rarely or never voting, and more likely to 
have a high-school education or less. People making $75,000 or more were more likely to be 
classified as always voting, as well as more likely to be college educated.

## Modeling

```{r}

income_votefreq_model <- lm(formula = voter_cat_coded ~ income_cat*educ,
                            data = voter_survey_538)

summary(income_votefreq_model)
```

Visualize effect of income and education on voter frequency.

```{r}
effect("income_cat*educ", income_votefreq_model) %>%
  data.frame()  %>%
  ggplot(aes(x = reorder(income_cat, fit),
             y = fit,
             ymin = lower,
             ymax = upper,
             label = format(fit, digits = 3))) +
  geom_errorbar() +
  geom_label() +
  facet_wrap(~reorder(educ, lower), ncol = 1) +
  labs(y = "Voter Category: 1-rarely/never, 2-sporadic, 3-always",
       x = "Income Category",
       title = "Effect of Income and Education on Voting Frequency",
       caption = "Data from FiveThirtyEight")
```

*Figure 8*

## Interpretation

Income and voting frequency have a weak (r-squared: .042) but statistically significant 
(p < 2.2e-16) positive relationship, even when education is taken into account. Lower 
income people are less likely to vote frequently and vice versa. However, education 
level has a noticeable effect on voting frequency, and that effect becomes more pronounced 
as income decreases. College-educated people with lower incomes are significantly more 
likely to vote more frequently than their non-college educated counterparts. However, as 
income level increases, education appears to affect voting frequency less. People of all 
education levels have similar likelihood of voting if they are high income. 

For context, it is important to note that people with higher income levels are more likely 
to have college educations. This is due to the high cost of higher education in the United 
States, which makes it far more accessible to wealthy people than to poor people. Additionally, 
people who live in low-income districts are less likely to receive a K-12 education that 
prepares them for college because school funding is based on local taxes. School districts 
where people have higher incomes and pay more taxes have more funding and better outcomes for 
their students, making it very difficult for poor districts to compete. 


# Results


# Further References

* Census Data: <https://www.census.gov/data/tables/time-series/demo/voting-and-registration/p20-585.html>
* Explanation of Census Voting Data: <https://www.census.gov/topics/public-sector/voting/about.html>
* FiveThirtyEight Data: <https://github.com/fivethirtyeight/data/tree/master/non-voters>
* Explanation of FiveThirtyEight Data: <https://github.com/fivethirtyeight/data/tree/master/non-voters#readme>
* Original FiveThirtyEight story: <https://projects.fivethirtyeight.com/non-voters-poll-2020-election/>
